Two subtle problems with over-representation analysis

### 2024-07-22 manuscript/manuscript.Rmd
grammar

To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
approach to a previously published dataset derived from 37 lung cancer patients, having tumor and normal
tissues for each patient [@Cho2022-ri].
These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
After reading them into R, we noticed some gene names converted to dates [@Abeysooriya2021-qt], so we
used the HGNChelper package to fix them [@Oh2022-gj].
DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
differentially expressed in tumor compared to normal tissue.
FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
ORA was conducted using fora, using genes with differential expression FDR<0.05.
Separate tests were conducted for up- and down-regulated genes.
For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
included.
Reactome pathways were used.
For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
We then calculated the number of pathways identified in downsampled data that were consistent with the full
dataset.
The number of consistent pathways gives an indication of sensitivity of each method.
Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
full dataset.
The proportion of significant pathways classified as inconsistent provides an estimate of the real false
discovery rate.
This was repeated 50 times at each sample size, and the median values were reported.

@@

To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
systematic downsampling of a gene expression dataset cosisting of normal and tumor samples
of 37 lung cancer patients.
In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
Fgsea yielded consistently more significant pathways than fora in downsampled data
(**Figure 4A**).
Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
compared to 41.5% for fora, confirming better recall for FCS over ORA.
It was noted that fgsea identified more pathways in downsampled data that were
inconsistent with the full dataset as compared to fora (**Figure 4B**).
However, inconsistent pathways were a smaller proportion of all findings for fgsea as
compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
compared to ORA methods.

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
The blue and red horizontal lines indicate the number of significant pathways identified
in the full set of 27 patients using fgsea and fora respectively.
(B) The number of significant pathways identified in downsampled data that are not
consistent the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 1 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.

>>>>

To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
approach to a previously published dataset derived from 37 lung cancer patients, having tumour and normal
tissues for each patient [@Cho2022-ri].
These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
After reading them into R, we noticed some gene names converted to dates [@Abeysooriya2021-qt], so we
used the HGNChelper package to fix them [@Oh2022-gj].
DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
differentially expressed in tumour compared to normal tissue.
FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
ORA was conducted using fora, using genes with differential expression FDR<0.05.
Separate tests were conducted for up- and down-regulated genes.
For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
included.
Reactome pathways were used.
For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
We then calculated the number of pathways identified in downsampled data that were consistent with the full
dataset.
The number of consistent pathways gives an indication of sensitivity of each method.
Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
full dataset.
The proportion of significant pathways classified as inconsistent provides an estimate of the real false
discovery rate.
This was repeated 50 times at each sample size, and the median values were reported.

@@

To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
systematic downsampling of a gene expression dataset consisting of normal and tumour samples
of 37 lung cancer patients.
In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
Fgsea yielded consistently more significant pathways than fora in downsampled data
(**Figure 4A**).
Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
compared to 41.5% for fora, confirming better recall for FCS over ORA.
It was noted that fgsea identified more pathways in downsampled data that were
inconsistent with the full dataset as compared to fora (**Figure 4B**).
However, inconsistent pathways were a smaller proportion of all findings for fgsea as
compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
compared to ORA methods.

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
The blue and red horizontal lines indicate the number of significant pathways identified
in the full set of 37 patients using fgsea and fora respectively.
(B) The number of significant pathways identified in downsampled data that are not
consistent with the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 1 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.
